DNA Methylation Hierarchical Variance Model\nStan Results

### 2020-10-01 StanAnalysis_updated.Rmd
differential methylation, clean up

stanFile = "FullResultsTCGA_relaxgamma3_postsum3.Rdata"
load(here(stanFile))
```

@@

<!-- \newpage -->
<!-- \section{Patient-specific analysis} -->
<!-- Basic overview: using a specified number of genes (here 100), take top genes (here using score_prob bootstrapping percentiles) to select a "gene signature" of sorts. Then for each patient, calculate the mean pair-wise distance within the individual patients' tumor samples. Method 1 is these raw values, method 2 normalizes by the mean PWD of all normal samples. Method 3 just examines the overall mean PWD for all tumor samples, normalized by mean PWD of all normal. -->

<!-- ```{r PatientPWD, echo=FALSE, message=FALSE, warning=FALSE, comment=NA, fig.height=9, fig.width=7} -->
<!-- library(gtools) -->

<!-- # pull out all samples beta values -->
<!-- all_beta <- FullAnnotation[,10:57] -->

<!-- # extract patient and tissue labels -->
<!-- patientLabel <- substr(colnames(all_beta), 1, 1) -->
<!-- patientLabel[10:12] <- "K*" -->
<!-- tissueLabel <- substr(colnames(all_beta), 2, 2) -->
<!-- tissueLabel[!(tissueLabel %in% "N")] <- "T"   #replace A/B/D/M with T -->

<!-- pats = patientLabel %>% sort %>% unique -->
<!-- npat = length(pats) -->

<!-- enrichedGenes <- permed_data %>% filter(score_prob_v3 < quantile(score_prob_v3, 0.05, na.rm = TRUE)) %>% -->
<!--   select(UCSC_RefGene_Name, score_prob, score_prob_v3, n) %>% -->
<!--   rename(Gene=UCSC_RefGene_Name, Pctile=score_prob_v3) -->

<!-- #ngene = 70 # using top N genes -->
<!-- #genes = enrichedGenes %>% top_n(-ngene,Pctile) %>% pull(Gene) -->
<!-- #genes = genes[1:ngene] -->

<!-- genes = enrichedGenes$Gene -->
<!-- ngene = length(genes) -->

<!-- pwd <- function(x) { -->
<!--   mean(dist(x)) -->
<!-- } -->
<!-- pwd2 <- function(x) { -->
<!--   abs(diff(x)) -->
<!-- } -->

<!-- signature_mat1 = matrix(,ngene,npat) # raw tumor pwd -->
<!-- signature_mat2 = matrix(,ngene,npat) # tumor pwd normalized by all normal pwd -->
<!-- for (igene in 1:ngene) { -->
<!--   # each gene, get gene inds, calculate PWD for the tumor samples -->
<!--   geneind = grep(genes[igene], FullAnnotation$UCSC_RefGene_Name) -->

<!--   # calculate PWD of each tumor -->
<!--   tum_pwd <- numeric(npat) -->
<!--   for (ipat in 1:npat) { -->
<!--     # each patient, get tumor sample inds -->
<!--     patind <- which(patientLabel == pats[ipat] & tissueLabel == "T") -->
<!--     tum_pwd[ipat] <- all_beta[geneind,patind] %>% logit %>% apply(1,pwd2) %>% mean -->
<!--   } -->

<!--   # calculate PWD of all normal -->
<!--   norm_pwd <- all_beta[geneind,tissueLabel == "N"] %>% logit %>% apply(1, pwd) %>% mean -->

<!--   signature_mat1[igene,] = tum_pwd -->
<!--   signature_mat2[igene,] = tum_pwd/norm_pwd -->
<!-- } -->

<!-- #cat("  \n\\newline  \nMethod 1 - raw intratumor PWD within each patient") -->
<!-- #signature_mat1 %>% as_tibble %>% `colnames<-`(pats) %>% gather %>% -->
<!-- #  mutate(gene=rep(genes,npat)) %>% rename(Patient=key,PWD=value) %>% -->
<!-- #  ggplot + geom_tile(aes(y=gene,x=Patient,fill=PWD)) + ylab(paste0(ngene," genes")) + -->
<!-- #  theme(axis.text.y = element_blank()) -->
<!-- #cat("  \n\\newline  \nMethod 2 - intratumor PWD normalized by all normal PWD") -->
<!-- #signature_mat2 %>% as_tibble %>% `colnames<-`(pats) %>% gather %>% -->
<!-- #  mutate(gene=rep(genes,npat)) %>% rename(Patient=key,PWD=value) %>% -->
<!-- #  ggplot + geom_tile(aes(y=gene,x=Patient,fill=PWD)) + ylab(paste0(ngene," genes")) + -->
<!-- #  theme(axis.text.y = element_blank()) -->
<!-- #cat("  \n\\newline  \nMethod 3 - all tumor PWD normalized by all normal PWD") -->
<!-- #signature_mat3 %>% as_tibble %>% mutate(gene=genes) %>% rename(PWD=V1) %>% -->
<!-- #  ggplot + geom_tile(aes(y=gene,x=1,fill=PWD)) + ylab(paste0(ngene," genes")) + -->
<!-- #  theme(axis.text.y = element_blank()) -->

<!-- library(pheatmap) -->
<!-- library(dendsort) -->

<!-- cat("  \n\\newline  \nMethod 1 - raw intratumor PWD within each patient") -->
<!-- df1 = as.data.frame(signature_mat1) -->
<!-- rownames(df1) <- genes -->
<!-- colnames(df1) <- pats -->
<!-- sort_hclust <- function(...) as.hclust(dendsort(as.dendrogram(...))) -->
<!-- mat_cluster_cols <- sort_hclust(hclust(dist(t(df1)))) -->
<!-- mat_cluster_rows <- sort_hclust(hclust(dist(df1))) -->
<!-- pheatmap( -->
<!--   mat=df1, -->
<!--   cluster_cols = mat_cluster_cols, -->
<!--   cluster_rows = mat_cluster_rows, -->
<!--   show_rownames = FALSE, -->
<!--   show_colnames = TRUE, -->
<!--   main = "Top 1118 Genes: Tumor PWD" -->
<!-- ) -->

<!-- cat("  \n\\newline  \nMethod 2 - intratumor PWD normalized by all normal PWD") -->
<!-- df1 = as.data.frame(signature_mat2) -->
<!-- rownames(df1) <- genes -->
<!-- colnames(df1) <- pats -->
<!-- sort_hclust <- function(...) as.hclust(dendsort(as.dendrogram(...))) -->
<!-- mat_cluster_cols <- sort_hclust(hclust(dist(t(df1)))) -->
<!-- mat_cluster_rows <- sort_hclust(hclust(dist(df1))) -->
<!-- pheatmap( -->
<!--   mat=df1, -->
<!--   cluster_cols = mat_cluster_cols, -->
<!--   cluster_rows = mat_cluster_rows, -->
<!--   show_rownames = FALSE, -->
<!--   show_colnames = TRUE, -->
<!--   main = "Top 1118 Genes: Tumor PWD" -->
<!-- ) -->
<!-- ``` -->

@@

enrichedPaths <- tibble(Pathway = enrich_sigmaP$Description, ngenes = enrich_sigmaP$Count, padj = enrich_sigmaP$p.adjust) %>% arrange(padj)

@@

enrichedPaths <- tibble(Pathway = enrich_sigmaPT$Description, ngenes = enrich_sigmaPT$Count, padj = enrich_sigmaPT$p.adjust) %>% arrange(padj)

@@

enrichedPaths <- tibble(Pathway = enrich_sigmaT$Description, ngenes = enrich_sigmaT$Count, padj = enrich_sigmaT$p.adjust) %>% arrange(padj)

@@

<!-- \newpage -->
<!-- \subsection{A4: Score Prob} -->
<!-- Genes are selected with cutoff >95th percentile (adjusted bootstrap), showing top 50 genes/pathways -->
<!-- ```{r App_score_prob, echo=FALSE, comment=NA, message=FALSE} -->
<!-- enrichedGenes <- permed_data %>%  -->
<!--   select(UCSC_RefGene_Name, score_prob, score_prob_v3, n) %>% -->
<!--   rename(Gene=UCSC_RefGene_Name, Score_median=score_prob, Pctile=score_prob_v3) %>% -->
<!--   filter(Pctile > quantile(Pctile, 0.95, na.rm = TRUE)) %>% -->
<!--   arrange(Pctile) -->

<!-- enrichedPaths <- tibble(Pathway = enrich_score_prob$Description, ngenes = enrich_score_prob$Count, padj = enrich_score_prob$p.adjust) %>% arrange(padj) -->

<!-- ngene <- dim(enrichedGenes)[1] -->
<!-- npath <- length(enrich_score_prob$Description) -->

<!-- cat(paste0("Total number of genes: ", ngene, "     Number of enrichment pathways: ", npath)) -->

<!-- nshow = 50 -->
<!-- if (ngene > nshow) { -->
<!--   enrichedGenes <- enrichedGenes[1:nshow,] -->
<!-- } -->
<!-- if (npath > nshow) { -->
<!--   enrichedPaths <- enrichedPaths[1:nshow,] -->
<!-- } -->

<!-- enrichedGenes %>% pander() -->

<!-- cat("Score_prob top 50 enriched paths:  \n") -->
<!-- enrichedPaths %>% pander() -->
<!-- ``` -->

<!-- \newpage -->
<!-- \subsection{A5: Score Mean} -->
<!-- Genes are selected with cutoff >95th percentile (adjusted bootstrap), showing top 50 genes/pathways -->
<!-- ```{r App_score_mean, echo=FALSE, comment=NA, message=FALSE} -->
<!-- enrichedGenes <- permed_data %>%  -->
<!--   select(UCSC_RefGene_Name, score_mean, score_mean_v3, n) %>% -->
<!--   rename(Gene=UCSC_RefGene_Name, Score_median=score_mean, Pctile=score_mean_v3) %>% -->
<!--   filter(Pctile > quantile(Pctile, 0.95, na.rm = TRUE)) %>% -->
<!--   arrange(Pctile) -->

<!-- enrichedPaths <- tibble(Pathway = enrich_score_mean$Description, ngenes = enrich_score_mean$Count, padj = enrich_score_mean$p.adjust) %>% arrange(padj) -->

<!-- ngene <- dim(enrichedGenes)[1] -->
<!-- npath <- length(enrich_score_mean$Description) -->

<!-- cat(paste0("Total number of genes: ", ngene, "     Number of enrichment pathways: ", npath)) -->

<!-- nshow = 50 -->
<!-- if (ngene > nshow) { -->
<!--   enrichedGenes <- enrichedGenes[1:nshow,] -->
<!-- } -->
<!-- if (npath > nshow) { -->
<!--   enrichedPaths <- enrichedPaths[1:nshow,] -->
<!-- } -->

<!-- enrichedGenes %>% pander() -->

<!-- cat("Score_mean top 50 enriched paths:  \n") -->
<!-- enrichedPaths %>% pander() -->
<!-- ``` -->

@@

\newpage
\subsection{A7: Score Median2}
Genes are selected with cutoff >95th percentile (adjusted bootstrap), showing top 50 genes/pathways
```{r App_score_median2, echo=FALSE, comment=NA, message=FALSE}
enrichedGenes <- permed_data %>% 
  select(UCSC_RefGene_Name, score_median2, score_median2_v3, n) %>%
  rename(Gene=UCSC_RefGene_Name, Score_median2=score_median2, Pctile=score_median2_v3) %>%
  filter(Pctile > quantile(Pctile, 0.95, na.rm = TRUE)) %>%
  arrange(Pctile)

enrichedPaths <- tibble(Pathway = enrich_score_median2$Description, ngenes = enrich_score_median2$Count, padj = enrich_score_median2$p.adjust) %>% arrange(padj)

ngene <- dim(enrichedGenes)[1]
npath <- length(enrich_score_median2$Description)

cat("Score_median2 top 50 enriched paths:  \n")
enrichedPaths %>% pander()
```

\newpage
\subsection{A9: Score Regularized}
Genes are selected with cutoff >95th percentile (adjusted bootstrap), showing top 50 genes/pathways
```{r App_score_reg, echo=FALSE, comment=NA, message=FALSE}
enrichedGenes <- permed_data %>% 
  select(UCSC_RefGene_Name, score_reg, score_reg_v3, n) %>%
  rename(Gene=UCSC_RefGene_Name, Score_reg=score_reg, Pctile=score_reg_v3) %>%
  filter(Pctile > quantile(Pctile, 0.95, na.rm = TRUE)) %>%
  arrange(Pctile)

enrichedPaths <- tibble(Pathway = enrich_score_reg$Description, ngenes = enrich_score_reg$Count, padj = enrich_score_reg$p.adjust) %>% arrange(padj)

ngene <- dim(enrichedGenes)[1]
npath <- length(enrich_score_reg$Description)

cat("Score_reg top 50 enriched paths:  \n")
enrichedPaths %>% pander()
```

>>>>

stanFile = "FullResultsTCGA_relaxgamma3.Rdata"
load(here(stanFile))
```

@@

@@

enrichedPaths <- tibble(Pathway = enrich_sigmaP$Description, ngenes = enrich_sigmaP$Count, padj = enrich_sigmaP$p.adjust) %>% arrange(padj)
geneList <- enrich_reactome$geneID %>% strsplit(split='/') %>%
  lapply(function(x) AnnotationDbi::mapIds(org.Hs.eg.db::org.Hs.eg.db, x, 'SYMBOL', 'ENTREZID')) %>%
  lapply(function(x) paste(x,collapse=";")) %>% unlist
enrichedPaths$Genes = geneList

@@

enrichedPaths <- tibble(Pathway = enrich_sigmaPT$Description, ngenes = enrich_sigmaPT$Count, padj = enrich_sigmaPT$p.adjust) %>% arrange(padj)
geneList <- enrich_reactome$geneID %>% strsplit(split='/') %>%
  lapply(function(x) AnnotationDbi::mapIds(org.Hs.eg.db::org.Hs.eg.db, x, 'SYMBOL', 'ENTREZID')) %>%
  lapply(function(x) paste(x,collapse=";")) %>% unlist
enrichedPaths$Genes = geneList

@@

enrichedPaths <- tibble(Pathway = enrich_sigmaT$Description, ngenes = enrich_sigmaT$Count, padj = enrich_sigmaT$p.adjust) %>% arrange(padj)
geneList <- enrich_reactome$geneID %>% strsplit(split='/') %>%
  lapply(function(x) AnnotationDbi::mapIds(org.Hs.eg.db::org.Hs.eg.db, x, 'SYMBOL', 'ENTREZID')) %>%
  lapply(function(x) paste(x,collapse=";")) %>% unlist
enrichedPaths$Genes = geneList

@@

@@
